---
title: "Creating interactive graphs"
author: "Lisa Nicvert and Romain Frelat"
date: "2026-04-21"
categories: [r, interactive graph, visualization, plotly, ggiraph]
image: ""
toc: true
draft: false
lightbox: true
code-overflow: scroll
---

## Introduction

Interactive graphs are getting popular and easy to make in R.  

But **when does your plot benefits from being interactive?**  

> Only consider interactive graphs if the output can be an html page. With interactive graphs, you will not be able to print the document easily nor receive comments inside the document you shared.  

Here is a quick comparison of the respective advantages of static vs interactive graphs.  

|      Static         |  Interactive                 |
|:--------------------|:-----------------------------|
| **publication**     | **exploration**              |
| printed document    | identifying outliers |
| getting feedback on draft  | large and complex graphs *              |


<br>

*N.B. Complex graphs (e.g. with many overlapping lines) will be more readable with interactive graphs for the data exploration stage. Yet, for finalized results, well-designed simple graphs are always more impactful.*  

**Interactive maps** are especially popular because users like to zoom on specific locations. However, this topic will not be covered in this post. If interested, have a look at the R-packages [leaflet](https://rstudio.github.io/leaflet/index.html), [mapview](https://r-spatial.github.io/mapview/) or [tmaps](https://r-tmap.github.io/tmap/). 



### Packages

In the following blog post, we will show in parallel three ways to produce interactive graphs:  

1. ggplot + plotly
2. native plotly
3. ggiraph

[plotly](https://plotly.com/r/) is a general javascript-based graphing library, which has extensions in [for several programming languages](https://plotly.com/graphing-libraries/). In R, there are two main ways to interact with plotly: via ggplot or via its R native syntax.

[ggiraph](https://davidgohel.github.io/ggiraph/) is an extension of ggplot implementing additional `geom_` functions which make the corresponding plot element interactive (e.g. `geom_point` in base ggplot -> `geom_point_interactive` in ggiraph). Interactive geometries comes with a `tooltip` aesthetic, which allows to control the information displayed when the element is hovered.

A comparison between plotly and ggiraph is also available [in the ggiraph book](https://www.ardata.fr/ggiraph-book/#how-does-it-compare-to-plotly).

We will use the Palmer's penguins dataset as illustrative example.

```{r}
#| label: 'packages'
#| eval: false
pak::pkg_install(c("ggplot2", "plotly", "ggiraph", "palmerpenguins", "ggalluvial", "dplyr"))
```

<br/>

### Set-up

```{r}
#| label: 'setup'
#| warning: false

suppressWarnings({
  library(palmerpenguins)
  library(ggplot2)
  library(plotly)
  library(ggiraph)
  library(ggalluvial)
  library(dplyr)
})
ggplot2::theme_set(ggplot2::theme_minimal())

head(penguins)
dim(penguins)
penguins$ID <- paste(
  1:nrow(penguins),
  penguins$year,
  penguins$species,
  penguins$sex,
  sep = "_"
)

# Remove missing values
no_na <- apply(penguins, 1, function(x) !any(is.na(x)))
penguins <- penguins[no_na,]
```



## Scatter plot


### Simple scatter plot
Let's visualize the body mass in function of the flipper length, colored by species.

:::: {.panel-tabset}
## ggplotly

```{r}
g1 <- ggplot(
  data = penguins,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = species)) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Penguin species"
  )

p1_gg <- ggplotly(g1)
style(p1_gg, hoverinfo = "text", text = penguins$ID)
```

## plotly

```{r}
p1_py <- plot_ly(penguins) |>
  add_markers(
    x = ~flipper_length_mm,
    y = ~body_mass_g,
    color = ~species,
    colors = c("darkorange", "purple", "cyan4"),
    text = ~ID,
    hoverinfo = "text"
  ) |>
  layout(
    xaxis = list(title = "Flipper length (mm)"),
    yaxis = list(title = "Body mass (g)"),
    legend = list(title = list(text = "Penguin species"))
  )

p1_py
```

## ggiraph

```{r}
g1_gi <- ggplot(
  data = penguins,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point_interactive(
    aes(
      color = species,
      tooltip = ID
    )
  ) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Penguin species"
  )

girafe(ggobj = g1_gi)
```

::: {.callout-tip title="Displaying the graph"}
ggiraph lets you choose how to render the graph:

- if wrapped in `girafe()`, it is an interactive element;
- else, it renders as a normal ggplot (as shown below).

```{r}
g1_gi
```
:::

::::

### Regression

::: {.panel-tabset}
## ggplotly

```{r}
g2 <- ggplot(
  data = penguins,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = species), alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Penguin species"
  )

p2_gg <- ggplotly(g2)
p2_gg
```

## plotly

It is slightly more complicated to show regression with native plotly functions. First you need to calculate the linear regression and its prediction.

```{r}
# calculate linear regression
lm1 <- lm(body_mass_g ~ flipper_length_mm, data = penguins)
# make a sequence of 1000 values along x range
xrange <- range(penguins$flipper_length_mm, na.rm = TRUE)
xseq <- seq(xrange[1], xrange[2], length.out = 1000)
# predicts values on xseq
predlm1 <- predict(lm1, data.frame(flipper_length_mm = xseq), se = TRUE)

# make the interactive graph
plot_ly(penguins) |>
  add_markers(
    x = ~flipper_length_mm,
    y = ~body_mass_g,
    color = ~species,
    colors = c("darkorange", "purple", "cyan4"),
    text = ~ID,
    hoverinfo = "text"
  ) |>
  add_ribbons(
    x = xseq,
    ymin = predlm1$fit - 1.96 * predlm1$se.fit,
    ymax = predlm1$fit + 1.96 * predlm1$se.fit,
    color = I("gray80"),
    showlegend = FALSE
  ) |>
  add_lines(
    x = xseq,
    y = predlm1$fit,
    color = I("steelblue"),
    showlegend = FALSE
  ) |>
  layout(
    xaxis = list(title = "Flipper length (mm)"),
    yaxis = list(title = "Body mass (g)"),
    legend = list(title = list(text = "Penguin species"))
  )
```

## ggiraph

In ggiraph, there is no native way to display information about different points of a line: only one element can be shown for the whole line (for instance, the model equation).

```{r}
# Precompute the model and save formula / R2
model <- lm(body_mass_g ~ flipper_length_mm, data = penguins)
coefs <- coef(model)
formula_label <- paste0(
  "y = ", round(coefs[2], 2), "x +", round(coefs[1], 2), "\n",
  "R² = ", round(summary(model)$r.squared, 2)
)

g2_gi <- ggplot(
  data = penguins,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point_interactive(
    aes(
      color = species, 
      tooltip = paste(
        "Flipper length (mm): ", flipper_length_mm, "\n",
        "Body mass (g): ", body_mass_g, "\n",
        "Species:", species
      )), 
      alpha = 0.8) +
  geom_smooth_interactive(
    method = "lm", 
    se = TRUE,
    aes(tooltip = formula_label)
    ) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Penguin species"
  )

girafe(g2_gi)
```
:::



## Boxplot

### Simple boxplot

::: {.panel-tabset}
## ggplotly

```{r}
g3 <- ggplot(
  penguins,
  aes(x = species, y = flipper_length_mm, color = species)
) +
  geom_boxplot(width = 0.3, show.legend = FALSE) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(x = "Species", y = "Flipper length (mm)")

p3_gg <- ggplotly(g3)
p3_gg
```


## plotly

```{r}
p3_py <- plot_ly(
  penguins,
  x = ~species,
  y = ~flipper_length_mm,
  color = ~species,
  colors = c("darkorange", "purple", "cyan4"),
  type = "box"
) |>
  layout(
    xaxis = list(title = "Species"),
    yaxis = list(title = "Flipper length (mm)")
  )

p3_py
```

## ggiraph

The information we want to display on the boxpluts is not present in the original data: to choose the `tooltip`, we use `after_stat` which lets us access these values. The available values can be inspected by calling `layer_data` on our graph.

```{r}
g3_gi <- ggplot(
  penguins,
  aes(x = species, y = flipper_length_mm, color = species)
) +
  geom_boxplot_interactive(
    width = 0.3, 
    show.legend = FALSE,
    aes(
      tooltip = after_stat({
        paste0(
          "species: ", .data$color,
          "\nq1: ", .data$lower,
          "\nmedian: ", .data$middle,
          "\nq3: ", .data$upper
        )
      })
    )
    ) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(x = "Species", y = "Flipper length (mm)")

# To see the column names we can use in after_stat, we can inspect layer_data():
# layer_data(g3_gi)
girafe(g3_gi)
```
:::


### Boxplot per group

::: {.panel-tabset}
## ggplotly

```{r}
g4 <- ggplot(
  penguins |> filter(!is.na(sex)),
  aes(x = species, y = flipper_length_mm, fill = sex)
) +
  geom_boxplot() +
  scale_fill_manual(values = c("darkorange", "cyan4")) +
  labs(x = "Species", y = "Flipper length (mm)", fill = "Sex")

p4_gg <- ggplotly(g4)
layout(p4_gg, boxmode = "group")
```


## plotly

```{r}
p4_py <- plot_ly(
  penguins,
  x = ~species,
  y = ~flipper_length_mm,
  color = ~sex,
  colors = c("darkorange", "cyan4"),
  type = "box"
) |>
  layout(
    xaxis = list(title = "Species"),
    yaxis = list(title = "Flipper length (mm)"),
    boxmode = "group"
  )

p4_py
```

## ggiraph

```{r}
g4_gi <- ggplot(
  penguins |> filter(!is.na(sex)),
  aes(x = species, y = flipper_length_mm, fill = sex)
) +
  geom_boxplot_interactive(
    aes(
      tooltip = 
        after_stat({
          paste0(
            "sex: ", .data$fill,
            "\nq1: ", .data$lower,
            "\nmedian: ", .data$middle,
            "\nq3: ", .data$upper
          )
        })
    )
  ) +
  scale_fill_manual(values = c("darkorange", "cyan4")) +
  labs(x = "Species", y = "Flipper length (mm)", fill = "Sex")

girafe(g4_gi)
```
:::


## Histogram and bars

### Histogram

::: {.panel-tabset}
## ggplotly

```{r}
g5 <- ggplot(data = penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_histogram(
    alpha = 0.5,
    position = "identity",
    binwidth = 2
  ) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(x = "Flipper length (mm)", y = "Frequency")

p5_gg <- ggplotly(g5)
p5_gg
```

## plotly

```{r}
p5_py <- plot_ly(
  penguins,
  x = ~flipper_length_mm,
  color = ~species,
  colors = c("darkorange", "purple", "cyan4"),
  type = "histogram",
  alpha = 0.5,
  xbins = list(size = 2)
) |>
  layout(
    xaxis = list(title = "Flipper length (mm)"),
    yaxis = list(title = "Frequency"),
    barmode = "overlay"
  )

p5_py
```

*N.B.: it is important to force bin size identical across species, else representation might be misleading, unless histnorm = "density".*

## ggiraph

```{r}
g5_gi <- ggplot(data = penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_histogram_interactive(
    alpha = 0.5,
    position = "identity",
    binwidth = 2,
    aes(tooltip = after_stat({
      paste0(
        "species: ", .data$fill,
        "\ncount: ", .data$count,
        "\nFlipper length: ", .data$xmin, "-", .data$xmax
        )
    })) 
  ) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(x = "flipper length (mm)", y = "Frequency")

girafe(g5_gi)
```
:::

### Barplot

::: {.panel-tabset}
## ggplotly

```{r}
g6 <- ggplot(
  data = penguins,
  aes(x = year, fill = species)
) +
  geom_bar() +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(
    x = "Year",
    y = "Count",
    fill = "Species"
  )
geom_bar(position = "dodge")
p6_gg <- ggplotly(g6)
layout(p6_gg, hovermode = "x unified")
```

## plotly

```{r}
year_species <- table(penguins$species, penguins$year) |> as.data.frame()
names(year_species) <- c("species", "year", "count")

p7_py <- plot_ly(
  year_species,
  x = ~year,
  y = ~count,
  color = ~species,
  colors = c("darkorange", "purple", "cyan4"),
  type = 'bar'
) |>
  layout(
    barmode = "stack",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Count"),
    legend = list(title = list(text = "Species")),
    hovermode = "x unified"
  )

p7_py
```

## ggiraph

```{r}
g6_gi <- ggplot(
  data = penguins,
  aes(x = year, fill = species)
) +
  geom_bar_interactive(aes(
    tooltip = after_stat({
      paste0(
        "species: ", .data$fill,
        "\ncount: ", .data$count,
        "\nyear: ", .data$x
      )
    })
  )) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(
    x = "Year",
    y = "Count",
    fill = "Species"
  )

girafe(g6_gi)
```
:::



## To go further

- palmerpenguins R package - [link](https://allisonhorst.github.io/palmerpenguins/)
- plotly [documentation](https://plotly.com/r/) and [book](https://plotly-r.com/) 
- ggiraph [documentation](https://davidgohel.github.io/ggiraph/index.html) and [book](https://www.ardata.fr/ggiraph-book/)
- ggplot documentation - [link](https://ggplot2-book.org/)


## Other tips and tricks

### Export as html

To share easily the interactive graphs in an html file, it might be easier to create self-contained HTML (even if the file size will be large, there will be only one single file).

In a quarto document, make sure to have in the header:
```{yaml}
format:
  html:
    embed-resources: true
```


To save a single plot in an self-contained html file, you can use: 
```{r}
#| eval: false
htmlwidgets::saveWidget(p1_py, "plot1.html")

htmlwidgets::saveWidget(girafe(g1_gi), "plot1_gi.html")
```

### Render large dataset
If you have many data points, use WebGL which is a lot more efficient at rendering heavy dataset.

```{r}
plotly::toWebGL(p1_py)
```

### Remove the long list of icons

::: {.panel-tabset}
## plotly
```{r}
p1_py |>
  config(
    modeBarButtons = list(list("toImage")),
    displaylogo = FALSE
  )
```

## ggiraph

By default, ggiraph displays only the saving and fullscreen icons (version 0.9.6). They can be removed as showed below (see [`opts_toolbar`](https://davidgohel.github.io/ggiraph/reference/opts_toolbar.html) documentation): 

```{r}
girafe(
  g1_gi, 
  options = list(
    opts_toolbar(
      saveaspng = FALSE, # don't show save PNG
      hidden = c("fullscreen") # hide fullscreen button
      )
  )
  )
```

More generally, the `girafe` function lets you customize interactive elements. See the [ggiraph documentation](https://www.ardata.fr/ggiraph-book/customize.html) for more information.
:::

## Other fancy graphs


### Facet

For facet plot, it is easier to use the native `ggplot` functions.

::: {.panel-tabset}
## ggplotly

```{r}
g8 <- ggplot(
  data = penguins,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = sex)) +
  scale_color_manual(values = c("darkorange", "cyan4"), na.translate = FALSE) +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Penguin sex"
  ) +
  facet_wrap(~species)

p8_gg <- ggplotly(g8)
p8_gg
```

## plotly

The function to arrange multiple plots in plotly is `plotly::subplot()`. It is very flexible and can accomodate all type of graphs. But in our case, the sub plots per species have to be created one by one. We use the function split and lapply to automate the plot creation.

```{r}
penguins_per_species <- split(penguins, ~species)
p8_py <- lapply(penguins_per_species, function(df) {
  plot_ly(df) |>
    add_markers(
      x = ~flipper_length_mm,
      y = ~body_mass_g,
      color = ~sex,
      colors = c("darkorange", "cyan4"),
      legendgroup = ~sex,
      showlegend = df$species[1] == names(penguins_per_species)[1]
    ) |>
    layout(
      xaxis = list(
        tickangle = -45,
        title = ifelse(
          df$species[1] == names(penguins_per_species)[2],
          "Flipper length (mm)",
          ""
        )
      )
    )
})

subplot(
  p8_py,
  nrows = 1,
  shareX = TRUE,
  shareY = TRUE,
  titleX = TRUE,
  titleY = TRUE
) |>
  layout(
    legend = list(title = list(text = "Penguin sex"))
  )
```

## ggiraph

```{r}
g8_gi <- ggplot(
  data = penguins,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point_interactive(
    aes(
      color = sex,
      tooltip = paste0(
        "species: ", species,
        "\nsex: ", sex,
        "\nbody mass: ", body_mass_g,
        "\nflipper length: ", flipper_length_mm
      )
      )) +
  scale_color_manual(values = c("darkorange", "cyan4"), na.translate = FALSE) +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Penguin sex"
  ) +
  facet_wrap(~species)

girafe(g8_gi)
```
:::

### Multiple regressions

It is also much simpler in ggplot

::: {.panel-tabset}

## ggplotly
```{r}
g9 <- ggplot(
  data = penguins,
  aes(x = bill_length_mm, y = bill_depth_mm)
) +
  geom_point(aes(color = species), alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, aes(color = species)) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Penguin species",
  )

p9_gg <- ggplotly(g9)
p9_gg
```

## ggiraph
```{r}
g9_gi <- ggplot(
  data = penguins,
  aes(x = bill_length_mm, y = bill_depth_mm)
) +
  geom_point_interactive(
    aes(
      color = species,
      tooltip = paste0(
        "bill length: ", bill_length_mm,
        "\nbill depth: ", bill_depth_mm,
        "\nspecies: ", species
      )
      ), 
      alpha = 0.8) +
  geom_smooth_interactive(
    method = "lm", se = FALSE, 
    aes(
      color = species,
      tooltip = paste0("species: ", species)
      )
  ) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Penguin species",
  )

girafe(g9_gi)
```
:::


### Violin plot

In this case, plotly is much easier

::: {.panel-tabset}

## plotly
```{r}
p10_py <- plot_ly(
  penguins,
  x = ~species,
  y = ~flipper_length_mm,
  color = ~sex,
  colors = c("darkorange", "cyan4"),
  type = "violin",
  box = list(visible = T)
) |>
  layout(
    xaxis = list(title = "Species"),
    yaxis = list(title = "Flipper length (mm)"),
    legend = list(title = list(text = 'Sex')),
    violinmode = "group"
  )

p10_py
```

## ggiraph

```{r}
g10_gi <- ggplot(
  penguins,
  aes(x = species, y = flipper_length_mm, fill = sex, color = sex)) +
  geom_violin_interactive(
    alpha = 0.2, 
    aes(
      tooltip = paste0(
        "species: ", species,
        "\nsex :", sex)
        )) +
  geom_boxplot_interactive(
    alpha = 0.2, 
    aes(
      tooltip = 
        after_stat({
          paste0(
            "sex: ", .data$fill,
            "\nq1: ", .data$lower,
            "\nmedian: ", .data$middle,
            "\nq3: ", .data$upper
          )
        })
    )) +
  scale_fill_manual(values = c("darkorange", "cyan4")) +
  scale_colour_manual(values = c("darkorange", "cyan4"))

girafe(g10_gi)
```
:::

### Parallel plot

In this case, plotly is much easier (or check out [ggalluvial](https://corybrunson.github.io/ggalluvial/articles/ggalluvial.html)).

::: {.panel-tabset}

## plotly
```{r}
# Create dimensions (one per categorical variable)
dims <- list(
  list(label = "Species", values = penguins$species),
  list(label = "Island", values = penguins$island),
  list(label = "Sex", values = penguins$sex),
  list(label = "Year", values = penguins$year)
)

p10_py <- plot_ly(
  type = "parcats",
  dimensions = dims,
  line = list(
    color = as.integer(penguins$species),
    colorscale = list(
      list(0, "#1f77b4"),
      list(0.5, "#ff7f0e"),
      list(1, "#2ca02c")
    )
  ),
  arrangement = "freeform",
  hoveron = "category"
)
p10_py

```


## ggplot

Unfortunately, the interactive geometries corresponding to ggalluvial don't exist yet...

```{r}
# First transform the data
penguins_alluvial <- penguins
penguins_alluvial$year <- factor(penguins_alluvial$year)
penguins_alluvial <- penguins_alluvial |> 
  dplyr::count(island, sex, year, species, .drop = FALSE)

g10_gi <- ggplot(
  data = penguins_alluvial,
  aes(
    axis1 = species, 
    axis2 = island, 
    axis3 = sex, 
    axis4 = year,
    y = n
    )) +
    geom_alluvium(
      aes(fill = species), 
      aes.bind = "alluvia",
      alpha = 0.8
      ) +
    geom_stratum(alpha = 0) +
    geom_text_interactive(
      stat = "stratum", 
      aes(
        label = after_stat(stratum),
        tooltip = after_stat(stratum)
        )) +
    scale_fill_manual(values = c("#1f77b4", "#2ca02c", "#ff7f0e"))

girafe(g10_gi)
```
:::